---
title: "Big Data and Uncertainty Quantification - Classification Project"
authors: "Dmytro Vaskov, Mariia Salivon"
date: "2026-02-02"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
install.packages("caret")
library(tidyverse)
library(caret)
library(MASS)
library(class)
library(rpart)
library(pROC)
library(knitr)
library(dplyr)
```

# 1. Introduction and Dataset Selection
This project analyzes ICE arrests data to predict whether cases result in a 
final deportation order. We implement and compare multiple classification 
methods including Logistic Regression, Linear Discriminant Analysis, k-Nearest Neighbors, and Classification Trees.

**Dataset:** ICE Arrests, Detention and Deportation Data from Kaggle  
**Source:** https://www.kaggle.com/datasets/bwandowando/ice-arrests-detention-and-deportation-data  
**Target Variable:** `final_order_yes_no` (binary classification: YES vs NO)
```{r load-data}
#Dataset:
ice_data <- read.csv("C:/Users/Dmytro Vaskov/Downloads/arrests-latest.csv")
cat("Dataset dimensions:", dim(ice_data)[1], "rows,", dim(ice_data)[2], 
    "columns\n")
head(ice_data, 5)
```

# 2. Exploratory Data Analysis

## 2.1 Target Variable Distribution
```{r target-distribution}
#Checking target variable with NA:
table(ice_data$final_order_yes_no, useNA = "always")
#Proportions:
prop.table(table(ice_data$final_order_yes_no))
#Target distribution without NA:
ice_clean <- ice_data %>% 
  filter(final_order_yes_no != "")
nrow(ice_clean)

ggplot(ice_clean, aes(x = final_order_yes_no, fill = final_order_yes_no)) +
  geom_bar() +
  geom_text(stat='count', aes(label=..count..), vjust=-0.5) +
  labs(title = "Distribution of Final Order Status",
       x = "Final Order",
       y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("NO" = "blue", "YES" = "red"))
```
**Findings:**
- 9,372 observations (2.5% of the total observations) have missing target values
and will be removed
- After cleaning: 60.8% YES, 36.7% NO -> reasonably balanced classes
- Total usable observations: 367,695

## 2.2 Data Structure
```{r predictor-overview}
#Structure of key variables:
ice_clean %>% 
  dplyr::select(apprehension_state, gender, apprehension_criminality, 
         citizenship_country, birth_year) %>% 
  summary()

#Checking missing values in key predictors:
colSums(is.na(ice_clean)) %>% 
  sort(decreasing = TRUE) %>% 
  head(10)
```
**Findings:**
- duplicate_likely:     3,621 missing  (should not be critical)
- birth_year:               1 missing  (minimal)
- All other variables:      0 missing  (no missing values -> great)
## 2.3 Demographic Analysis
```{r demographics}
#Age variable
ice_clean <- ice_clean %>%
  mutate(
    apprehension_year = as.numeric(format(as.Date(apprehension_date), "%Y")),
    age = apprehension_year - birth_year
  )
summary(ice_clean$age)
#Removing unrealistic ages
ice_clean <- ice_clean %>%
  filter(age >= 0 & age <= 115)

#Age distribution
ggplot(ice_clean, aes(x = age)) +
  geom_histogram(bins = 50, fill = "blue", alpha = 0.7) +
  labs(title = "Age Distribution of Arrested Individuals",
       subtitle = "Age calculated at time of apprehension",
       x = "Age",
       y = "Count") +
  theme_minimal()
```
**Findings:**
- The distribution of age shows a right-skewed distribution with a peak at the
age of approximately 30. 
- The bell shape suggests the near normality, especially in the middle of the
distribution. 
- Very few cases for the people with the age below 20.
- The right tail is extended on the right side, which indicates that either the 
dataset is focused on the older group of people or that ICE enforcement 
activities predominantly target the working-age population, likely reflecting 
immigration and labor force demographics.
- Overall the data seems to be well suited for the statistical modeling with 
regressions and classifications, which assumes the normality of the predictors.

## 2.4 Gender Distribution
```{r gender-analysis}
#Final order & gender:
table(ice_clean$gender, ice_clean$final_order_yes_no)
ggplot(ice_clean, aes(x = gender, fill = final_order_yes_no)) +
  geom_bar(positions = "fill") +
  labs(title = "Final Order by Gender",
       x = "Gender",
       y = "Final order") +
  theme_minimal() +
  scale_fill_manual(values = c("NO" = "blue", "YES" = "red"))
```
**Findings:**
**Key Findings:**
- The dataset is heavily skewed toward male arrests, with males representing
approximately 85-90% of all cases, while females account for only 10-15%
- This extreme gender imbalance reflects the demographics of immigration 
enforcement, likely due to higher rates of male migration for labor-related 
reasons. 
- Very few cases have "Unknown" gender, indicating good data quality 
for this variable. 
- When examining the proportion plot, both males and females
show similar proportions of final order outcomes. 
- The fact that the blue-to-red ratio (NO vs YES) is consistent across both 
Male and Female categories suggests that gender alone may not be a strong 
discriminator for predicting final order status. The "Unknown" category is 
negligible and can likely be removed or merged during preprocessing without
significant impact

**Implication for Modeling:**
While gender should be included as a predictor for completeness, the similar 
outcome proportions across gender categories suggest it may have low 
individual predictive importance. However, it could still contribute to model 
performance when combined with other predictors in multivariate models.

## 2.5 Criminality Categories:
``` {r criminality}
table(ice_clean$apprehension_criminality)
ggplot(ice_clean, aes(x = apprehension_criminality, fill = final_order_yes_no))+
  geom_bar(position = "fill") +
  labs(title = "Final Order by Criminality Category",
       x = "Criminality Category",
       y = "Proportion",
       fill = "Final Order") +
  theme_minimal() +
  coord_flip() +
  scale_fill_manual(values = c("NO" = "blue", "YES" = "red"))
```

**Key Findings:**
- Criminality category is a strong predictor of final order outcomes, showing 
clear and substantial differences across categories. 
- Convicted criminals have the highest rate of final orders, indicating that 
individuals with criminal convictions are most likely to receive deportation orders. 
- Pending criminal charges show moderate final order rates, slightly below those
with convictions but still above the overall average. 
- Other immigration violators (no criminal background) have the lowest rate of
final order, suggesting approximately equal chances of receiving or not receiving
a final order. 
- The likelihood of receiving a final order increases with criminal severity: 
Other Immigration Violator < Pending Charges < Convicted Criminal
- The large red area for "1 Convicted Criminal" shows this is the most decisive
factor for deportation outcomes

**Implication for Modeling:**
Criminality category will likely be one of the most important predictors in our 
classification models. The clear separation between categories suggests:
- High discriminative power for all classification methods
- This variable should contribute significantly to model accuracy
- May be selected as a primary split variable in classification trees
- Strong evidence for including this as a key predictor in logistic regression 
and discriminant analysis models

## 2.6 Top Citizenship Countries
```{r citizenship}
#Top 10 citizenship countries
top_countries <- ice_clean %>%
  count(citizenship_country, sort = TRUE) %>%
  head(10)
print(top_countries)

ice_clean %>%
  filter(citizenship_country %in% top_countries$citizenship_country) %>%
  ggplot(aes(x = reorder(citizenship_country, citizenship_country, function(x) -length(x)), 
             fill = final_order_yes_no)) +
  geom_bar(position = "fill") +
  labs(title = "Final Order Status by Top 10 Citizenship Countries",
       x = "Country",
       y = "Proportion",
       fill = "Final Order") +
  theme_minimal() +
  coord_flip() +
  scale_fill_manual(values = c("NO" = "blue", "YES" = "red"))
```
**Key Findings:**
- Mexico dominates the dataset as the most represented country, which is 
expected given geographical proximity to the United States. While most countries
show 65-75% final order rates, there are notable differences:
- Higher final order rates (~75-85%): Mexico, Guatemala, Honduras, El Salvador 
(Central American countries)
- Moderate rates (~60-70%): Venezuela, Nicaragua, Colombia, Cuba, Dominican Republic
- Lower rates (~55-65%): Cuba, Ecuador (showing more balanced outcomes)
  
- Regional patterns emerge: Central American countries (Northern Triangle: 
Guatemala, Honduras, El Salvador + Mexico) tend to have higher final order rates
compared to South American and Caribbean nations.
- Cuba and Nicaragua show relatively lower final order rates compared to Central
American countries

**Possible explanations for variation:**
- Different bilateral agreements or immigration policies with specific countries
- Geographic proximity and ease of deportation logistics
- Varying rates of criminal involvement across nationalities
- Different asylum claim success rates by country of origin
- Political considerations in US-country relations

**Implication for Modeling:**
Citizenship country shows moderate predictive power with clear variation across
countries. However, with many countries represented in the full dataset, we will
need to:
- Group less frequent countries into "Other" category during preprocessing
- Consider this variable in multivariate models where it may interact with other predictors
- Use this as a categorical predictor in all classification models

# 3. Data Preprocessing
## 3.1 Data Cleaning
```{r data-cleaning}

#Removing the row with missing birth_year
ice_clean <- ice_clean %>%
  filter(!is.na(age))
ice_clean$final_order_yes_no <- factor(ice_clean$final_order_yes_no, 
                                        levels = c("NO", "YES"))

cat("Final dataset size:", nrow(ice_clean), "observations\n")
cat("Target distribution:\n")
table(ice_clean$final_order_yes_no)
```
## 3.2 Feature Selection
```{r feature-selection}
#Selecting relevant predictors for modeling (variables that are likely to predict
#final order status)
model_data <- ice_clean %>%
  dplyr::select(
    final_order_yes_no,
    age,
    gender,
    citizenship_country,
    apprehension_state,
    apprehension_criminality,
    apprehension_method,
    case_category,
    final_program
  )

#Handling empty strings
model_data <- model_data %>%
  mutate(
    apprehension_state = ifelse(apprehension_state == "", "Unknown", apprehension_state),
    citizenship_country = ifelse(citizenship_country == "", "Unknown", citizenship_country),
    case_category = ifelse(case_category == "", "Unknown", case_category)
  )

model_data <- model_data %>%
  mutate(across(where(is.character), as.factor))

#Checking structure
str(model_data)
summary(model_data)
```
## 3.3 Handling High-Cardinality Variables
```{r handle-cardinality}
#Some variables have too many categories, we have grouped less frequent ones
#into "Other"

#Group rare categories
group_rare_categories <- function(x, threshold = 0.01, other_label = "Other") {
  freq <- prop.table(table(x))
  rare <- names(freq[freq < threshold])
  levels(x)[levels(x) %in% rare] <- other_label
  return(x)
}

model_data <- model_data %>%
  mutate(
    citizenship_country = group_rare_categories(citizenship_country, threshold = 0.01),
    apprehension_state = group_rare_categories(apprehension_state, threshold = 0.01),
    case_category = group_rare_categories(case_category, threshold = 0.01)
  )

#New levels
cat("Citizenship countries (grouped):", nlevels(model_data$citizenship_country), "\n")
cat("Apprehension states (grouped):", nlevels(model_data$apprehension_state), "\n")
cat("Case categories (grouped):", nlevels(model_data$case_category), "\n")
```
## 3.4 Train-Test Split
```{r train-test-split}
set.seed(123) 
#Creating 70-30 split
train_index <- createDataPartition(model_data$final_order_yes_no, 
                                   p = 0.7, 
                                   list = FALSE)

train_data <- model_data[train_index, ]
test_data <- model_data[-train_index, ]

cat("Training set size:", nrow(train_data), "\n")
cat("Test set size:", nrow(test_data), "\n")
cat("\nTraining set target distribution:\n")
prop.table(table(train_data$final_order_yes_no))
cat("\nTest set target distribution:\n")
prop.table(table(test_data$final_order_yes_no))
```
## 3.5 Data Standardization (for LDA, QDA, kNN)
```{r standardization}
# For methods that require standardized data (LDA, QDA, kNN)
# We'll standardize the numeric predictor (age)
train_data$age <- as.numeric(train_data$age)
test_data$age <- as.numeric(test_data$age)

#Mean and sd from training data
age_mean <- mean(train_data$age, na.rm = TRUE)
age_sd <- sd(train_data$age, na.rm = TRUE)

#Copies for scaled versions
train_data_scaled <- train_data
test_data_scaled <- test_data

# Standardizing
train_data_scaled$age <- as.numeric((train_data$age - age_mean) / age_sd)
test_data_scaled$age <- as.numeric((test_data$age - age_mean) / age_sd)

#Verifying standartization
cat("Training set:\n")
cat("  Mean:", round(mean(train_data_scaled$age, na.rm = TRUE), 10), "\n")
cat("  SD:", round(sd(train_data_scaled$age, na.rm = TRUE), 4), "\n\n")
cat("Test set:\n")
cat("  Mean:", round(mean(test_data_scaled$age, na.rm = TRUE), 4), "\n")
cat("  SD:", round(sd(test_data_scaled$age, na.rm = TRUE), 4), "\n\n")

cat("Training observations:", nrow(train_data_scaled), "\n")
cat("Test observations:", nrow(test_data_scaled), "\n")
```

# 4. Model Implementation
We will implement and compare six classification methods:
1. Logistic Regression
2. Linear Discriminant Analysis (LDA)
3. Quadratic Discriminant Analysis (QDA)
4. k-Nearest Neighbors (k-NN)
5. Classification Tree (CART)
6. Random Forest

## 4.1 Logistic Regression
```{r logistic-regression}
#Logistic regression model
logit_model <- glm(final_order_yes_no ~ ., 
                   data = train_data, 
                   family = binomial)


#Predictions on test set
logit_probs <- predict(logit_model, test_data, type = "response")
logit_pred <- ifelse(logit_probs > 0.5, "YES", "NO")
logit_pred <- factor(logit_pred, levels = c("NO", "YES"))

# Confusion matrix
logit_cm <- confusionMatrix(logit_pred, test_data$final_order_yes_no)
print(logit_cm)

logit_accuracy <- logit_cm$overall['Accuracy']
```

**Key Findings from Logistic Regression:**
```{r logit-interpretation}
#Significant predictors
coef_summary <- summary(logit_model)$coefficients
significant_vars <- coef_summary[coef_summary[,4] < 0.001, ]
cat("\nMost significant predictors (p < 0.001):\n")
print(head(significant_vars[order(abs(significant_vars[,3]), decreasing = TRUE), ], 10))
```

**Key Findings:**
**The most influential categories are:**
- Reinstated Final Order (z = 84.5): Cases with previously issued orders that are being reinstated have the highest likelihood of resulting in deportation.
- Expedited Removal categories (z = 61.5): Cases processed through expedited 
removal pathways show very high deportation likelihood, reflecting the 
streamlined nature of these administrative processes.
- Administrative Final Orders (z = 83.1, 62.4): Cases that already have administratively final orders show strong positive associations, as expected 
since these represent completed enforcement actions.

These findings suggest that administrative pathway and case processing type are
the strongest determinants of the outcome.

**Secondary influential categories:**
Apprehension criminality shows highly significant effects, but interestingly with negative coefficients:
- Other Immigration Violator (beta = -0.85, z = -32.7): Individuals with no criminal background are significantly LESS likely to receive final orders compared to
convicted criminals (the reference category).
- Pending Criminal Charges (beta = -0.47, z = -20.5): Those with pending charges 
fall between convicted criminals and non-criminals in deportation likelihood.

This pattern confirms our EDA findings: criminal conviction status is a major
driver of deportation outcomes, with a clear hierarchy where convicted criminals
face the highest deportation rates.


```{r logit-interpretation}
#Odds ratios for interpretation
cat("\n\nOdds Ratios for Top Predictors:\n")
odds_ratios <- exp(coef(logit_model))
top_odds <- sort(odds_ratios[2:min(20, length(odds_ratios))], decreasing = TRUE)
print(head(top_odds, 10))
```
**Key Findings:**
The odds ratios reveal practical effect sizes:
- Arkansas (OR = 1.96): Cases in Arkansas are nearly twice as likely to result 
in deportation compared to the reference state, controlling for all other 
factors.
- Arizona (OR = 1.61) and California (OR = 1.52): Border states show elevated deportation odds, possibly reflecting proximity to removal infrastructure and enforcement priorities.
- Nicaragua (OR = 1.32), Honduras (OR = 1.25), Guatemala (OR = 1.19): Certain nationalities face higher deportation odds, potentially due to diplomatic 
agreements facilitating returns or lower asylum approval rates.
- Gender effects are minimal (Male OR = 1.06, Unknown OR = 1.45): Gender shows
weak effects once other factors are controlled, confirming it's not a primary
driver.

## 4.2 Linear Discriminant Analysis (LDA)
```{r lda-model}
# Fit LDA
lda_model <- lda(final_order_yes_no ~ ., data = train_data_scaled)
print(lda_model)

#Predictions
lda_pred <- predict(lda_model, test_data_scaled)
lda_class <- lda_pred$class

# Confusion matrix
lda_cm <- confusionMatrix(lda_class, test_data_scaled$final_order_yes_no)
print(lda_cm)

lda_accuracy <- lda_cm$overall['Accuracy']

cat("\nLDA Posterior Probability Summary:\n")
summary(lda_pred$posterior[,2])
```
**Key Findings:**
The LDA model achieves strong overall performance with 93.2% accuracy, 
substantially outperforming the no-information-rate baseline (62.3%). 
This demonstrates that linear discriminant analysis effectively separates the two 
classes using linear decision boundaries.

Key Performance Metrics:
- Specificity (98.3%) is very high. The model correctly identifies "YES" cases 
at a very high rate, with only 1,189 false negatives out of 68,759 actual 
YES cases.
- Sensitivity (84.8%) is also goof but lower. The model correctly identifies "NO"
cases 85% of the time, but misclassifies 6,330 NO cases as YES.
- Kappa (0.85) indicates excellent agreement beyond chance, confirming 
the model's strong discriminative ability.

Interpretation of the Trade-off:
The model shows a bias toward predicting deportation (YES).
This conservative approach means:
- High confidence in deportation predictions (96.7% precision for YES 
predictions)
- Some non-deportation cases incorrectly flagged as deportations

Posterior Probability Distribution:
The posterior probabilities reveal strong bimodal separation:
- Median posterior probability for YES class: 0.999 (extremely confident)
- 1st Quartile: 0.0002 vs 3rd Quartile: 0.9995

The linear discriminant function creates a clean decision boundary that
separates the vast majority of cases decisively.


## 4.3 Quadratic Discriminant Analysis (QDA)
Quadratic Discriminant Analysis could not be fitted due to rank deficiency - 
the high dimensionality of the predictor space (many categorical variables) 
relative to sample sizes in certain groups makes separate covariance matrix 
estimation infeasible. We proceed with LDA which assumes equal covariances 
across groups.

## 4.4 k-Nearest Neighbors (k-NN)
k-NN was excluded from this analysis due to computational constraints.
With 367,000+ observations and high-dimensional data (100+ features after 
encoding categorical variables), k-NN becomes computationally prohibitive.
The distance calculations required for k-NN scale poorly with both sample size 
and dimensionality. We proceed with the remaining five classification methods.

## 4.5 Classification Tree (CART)
```{r cart-model}
library(rpart.plot)

#Fit classification tree
tree_model <- rpart(final_order_yes_no ~ ., 
                    data = train_data,
                    method = "class",
                    control = rpart.control(cp = 0.001))

rpart.plot(tree_model, main = "Classification Tree for Final Order Prediction")

print(tree_model)

cat("\nVariable Importance:\n")
print(tree_model$variable.importance)

#Predictions
tree_pred <- predict(tree_model, test_data, type = "class")

#Confusion matrix
tree_cm <- confusionMatrix(tree_pred, test_data$final_order_yes_no)
print(tree_cm)

tree_accuracy <- tree_cm$overall['Accuracy']

#Pruning the tree using cross-validation
plotcp(tree_model)
cat("\nCross-validation results for tree complexity:\n")
printcp(tree_model)

#Optimal CP
optimal_cp <- tree_model$cptable[which.min(tree_model$cptable[,"xerror"]),"CP"]
cat("\nOptimal CP:", optimal_cp, "\n")

#Pruning
tree_pruned <- prune(tree_model, cp = optimal_cp)
rpart.plot(tree_pruned, main = "Pruned Classification Tree")

#Predictions with pruned tree
tree_pruned_pred <- predict(tree_pruned, test_data, type = "class")
tree_pruned_cm <- confusionMatrix(tree_pruned_pred, test_data$final_order_yes_no)
print(tree_pruned_cm)

tree_pruned_accuracy <- tree_pruned_cm$overall['Accuracy']
```

**Key Findings:**
The tree confirms that case category dominates all other predictors, making it 
the first and most important split. Administrative pathway determines outcome 
more than individual characteristics. The pruned tree is virtually identical to
the full tree, showing the model found a stable decision structure without 
unnecessary complexity.

## 4.6 Random Forest
```{r random-forest}
library(randomForest)

#Fit Random Forest
set.seed(123)
rf_model <- randomForest(final_order_yes_no ~ ., 
                         data = train_data,
                         ntree = 100,
                         mtry = sqrt(ncol(train_data) - 1),
                         importance = TRUE)
print(rf_model)
varImpPlot(rf_model, main = "Variable Importance in Random Forest")

#Importance scores
importance_scores <- importance(rf_model)
cat("\nTop 10 Most Important Variables:\n")
print(importance_scores[order(importance_scores[,3], decreasing = TRUE)[1:10], ])

#Predictions
rf_pred <- predict(rf_model, test_data)

#Confusion matrix
rf_cm <- confusionMatrix(rf_pred, test_data$final_order_yes_no)
print(rf_cm)

rf_accuracy <- rf_cm$overall['Accuracy']

# Plot error rate vs number of trees
plot(rf_model, main = "Random Forest Error Rate vs Number of Trees")
```
**Key Findings:**
**Variable Importance:**
Case category is the supreme predictor by both metrics, confirming that 
administrative processing pathway determines outcomes more than any individual characteristic. Criminality ranks high on Gini but lower on accuracy (less 
unique information beyond case category). Demographics (age, gender) contribute 
minimally, reinforcing that systemic/procedural factors outweigh personal 
attributes.
**Random Forest Error Rate vs Number of Trees:**
The plot shows error rates stabilizing very quickly:
- Black line (Overall Error ~6%): Total misclassification rate plateaus after 
just ~50 trees
- Red line (Class-specific Error ~2%): Error for one class stabilizes even faster
- Green line (Class-specific Error ~14%): Error for the other class also 
stabilizes early
Key Insight: 
Error rates converge by 50-100 trees and remain flat afterwards. This confirms 
that 100 trees is sufficient.
Thus, the Random Forest model is stable and not overfitting.

# 5. Model Comparison
## 5.1 Overall Performance Summary
```{r model-comparison}
#Comparison dataframe
model_comparison <- data.frame(
  Model = c("Logistic Regression", "LDA", 
            "Classification Tree", "Pruned Tree", "Random Forest"),
  Accuracy = c(logit_accuracy, lda_accuracy,
               tree_accuracy, tree_pruned_accuracy, rf_accuracy),
  Sensitivity = c(logit_cm$byClass['Sensitivity'],
                  lda_cm$byClass['Sensitivity'],
                  tree_cm$byClass['Sensitivity'],
                  tree_pruned_cm$byClass['Sensitivity'],
                  rf_cm$byClass['Sensitivity']),
  Specificity = c(logit_cm$byClass['Specificity'],
                  lda_cm$byClass['Specificity'],
                  tree_cm$byClass['Specificity'],
                  tree_pruned_cm$byClass['Specificity'],
                  rf_cm$byClass['Specificity']),
  F1_Score = c(logit_cm$byClass['F1'],
               lda_cm$byClass['F1'],
               tree_cm$byClass['F1'],
               tree_pruned_cm$byClass['F1'],
               rf_cm$byClass['F1'])
)

model_comparison[,2:5] <- round(model_comparison[,2:5], 4)
kable(model_comparison, 
      caption = "Performance Comparison of Classification Models",
      align = 'lcccc')

#Best model by each metric:
cat("\nBest Model by Metric:\n")
cat("Accuracy:", model_comparison$Model[which.max(model_comparison$Accuracy)], 
    "(", max(model_comparison$Accuracy), ")\n")
cat("Sensitivity:", model_comparison$Model[which.max(model_comparison$Sensitivity)], 
    "(", max(model_comparison$Sensitivity), ")\n")
cat("Specificity:", model_comparison$Model[which.max(model_comparison$Specificity)], 
    "(", max(model_comparison$Specificity), ")\n")
cat("F1 Score:", model_comparison$Model[which.max(model_comparison$F1_Score)], 
    "(", max(model_comparison$F1_Score), ")\n")
```
**Key Findings:**
Random Forest is the overall winner with highest accuracy (93.6%), sensitivity 
(84.9%), and F1 score (90.9%). Classification Tree achieved best specificity 
(99.3%) but lower overall accuracy. Logistic Regression and LDA both performed 
well (~93%) with better interpretability.

## 5.2 Accuracy Comparison Visualization
```{r accuracy-plot}
#Bar plot of accuracy
ggplot(model_comparison, aes(x = reorder(Model, Accuracy), y = Accuracy, fill = Model)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_text(aes(label = round(Accuracy, 4)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  ylim(0, 1) +
  labs(title = "Model Accuracy Comparison",
       x = "Model",
       y = "Accuracy") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10))
```

## 5.3 Sensitivity vs Specificity Trade-off
```{r sens-spec-plot}
#Scatter plot showing sensitivity-specificity trade-off
ggplot(model_comparison, aes(x = Specificity, y = Sensitivity, color = Model)) +
  geom_point(size = 4) +
  geom_text(aes(label = Model), vjust = -1, size = 3) +
  xlim(0, 1) +
  ylim(0, 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray") +
  labs(title = "Sensitivity vs Specificity Trade-off",
       subtitle = "Points closer to top-right corner are better",
       x = "Specificity (True Negative Rate)",
       y = "Sensitivity (True Positive Rate)") +
  theme_minimal()
```
**Key Findings:**
All five models cluster tightly in the top-right corner (high sensitivity ~85%,
very high specificity ~98-99%), indicating excellent overall performance with 
minimal trade-offs.
Key Observations:
- All models are nearly overlapping. They achieve very similar 
sensitivity-specificity balance.
The clustering shows that different modeling approaches converge on the same performance region, suggesting the data has clear separability between classes.
No model sacrifices one metric for the other. All achieve both high sensitivity
and specificity simultaneously, which is ideal for this classification problem.


## 5.4 Confusion Matrices Comparison
```{r confusion-matrices}
#Visual comparison of confusion matrices for top 3 models
plot_confusion_matrix <- function(cm, title) {
  cm_table <- as.data.frame(cm$table)
  ggplot(cm_table, aes(x = Reference, y = Prediction, fill = Freq)) +
    geom_tile() +
    geom_text(aes(label = Freq), size = 6) +
    scale_fill_gradient(low = "white", high = "steelblue") +
    labs(title = title,
         x = "Actual",
         y = "Predicted") +
    theme_minimal() +
    theme(legend.position = "none")
}

#Confusion matrices for best models
library(gridExtra)
p1 <- plot_confusion_matrix(rf_cm, "Random Forest")
p2 <- plot_confusion_matrix(logit_cm, "Logistic Regression")
p3 <- plot_confusion_matrix(lda_cm, "LDA")
p4 <- plot_confusion_matrix(tree_pruned_cm, "Pruned Tree")

grid.arrange(p1, p2, p3, p4, ncol = 2)
```
**Key Findings:**
All four models show strong performance with similar error patterns:
- Random Forest: Most balanced errors (6,284 false positives, 
744 false negatives) -> best overall
- Logistic Regression: Slightly more false positives (6,605) and more false 
negatives (880) -> still strong
- LDA: Highest false positives (6,330) but fewer false negatives (1,189)
- Pruned Tree: Highest false positives (6,716) but lowest false 
negatives (454) -> most conservative, strongly biased toward predicting YES
Pattern: All models make more false positive errors than false negatives. This conservative bias means they're better at catching actual deportation cases but sometimes over-predict deportations.
Winner: Random Forest has the best balance with fewest total errors.

## 5.5 Statistical Significance Testing
```{r mcnemar-test}
#McNemar's test to compare models
#Random Forest vs Logistic Regression
rf_correct <- (rf_pred == test_data$final_order_yes_no)
logit_correct <- (logit_pred == test_data$final_order_yes_no)

contingency <- table(rf_correct, logit_correct)
cat("\nContingency Table: Random Forest vs Logistic Regression\n")
print(contingency)

mcnemar_result <- mcnemar.test(contingency)
cat("\nMcNemar's Test Result:\n")
print(mcnemar_result)

if(mcnemar_result$p.value < 0.05) {
  cat("\nConclusion: The difference in performance between Random Forest and Logistic Regression is statistically significant (p < 0.05)\n")
} else {
  cat("\nConclusion: No statistically significant difference between Random Forest and Logistic Regression (p >= 0.05)\n")
}
```


# 6. Conclusion
## 6.1 Summary of Findings
This project successfully applied multiple advanced classification techniques 
to predict final deportation order outcomes using ICE arrest data. 
Through systematic exploratory data analysis, preprocessing, and model 
implementation, we achieved high predictive accuracy while gaining valuable 
insights into the factors that influence deportation decisions.
### Dataset Characteristics
The ICE arrests dataset provided a robust foundation for classification 
modeling with 367,695 usable observations after cleaning. The target variable 
(final deportation order: YES/NO) showed reasonable class balance 
(62% YES, 38% NO), which facilitated effective model training.
The dataset exhibited high quality with minimal missing values, though 
approximately 15% of observations had unknown apprehension states.
### Key Predictive Variables
**Strong Predictors:**
- Case category is the supreme predictor, demonstrating that administrative 
processing pathways determine outcomes more strongly than individual 
characteristics.
- Criminality status showed the clearest gradient: convicted criminals
had ~85-90% deportation rates, pending charges ~65-70%, and other immigration
violators ~50-55%, confirming criminal history as a decisive factor.

**Moderate Predictors:**
- Citizenship country exhibited variation across nations, with Central 
American countries (Mexico, Guatemala, Honduras) showing higher deportation 
rates (75-85%) compared to South American and Caribbean nations (55-70%).
- Age followed a right-skewed distribution peaking at 30-35 years, with 
approximate normality supporting parametric modeling assumptions.

**Weak Predictors:**
- Gender showed minimal predictive power, with similar outcome proportions 
across male and female categories despite the dataset's 85% male composition

### Model Performance Comparison

We implemented and compared five classification methods, each offering unique strengths:

**Random Forest (Best Overall Performance):**
- Accuracy: 93.6%
- Sensitivity: 84.9% 
- Specificity: 98.7%
- F1 Score: 90.9%
- Random Forest achieved the highest accuracy and best balance between 
sensitivity and specificity. Variable importance analysis confirmed case 
category as the dominant predictor, with criminality as a strong secondary 
factor. The error rate stabilization plot showed convergence after just 
50-100 trees, validating our choice of 100 trees and demonstrating efficient 
learning without overfitting.

**Logistic Regression (Best Interpretability):**
- Accuracy: 93.2%
- Sensitivity: 82.5%
- Specificity: 98.4%
- F1 Score: 89.8%
- Logistic regression provided strong performance with the added benefit of 
coefficient interpretability. This model is ideal to understand the magnitude 
and direction of predictor effects on deportation likelihood.

**Linear Discriminant Analysis (LDA):**
- Accuracy: 93.0%
- Sensitivity: 79.8%
- Specificity: 98.3%
- F1 Score: 88.5%
- LDA performed comparably to logistic regression while making explicit 
normality assumptions. The age distribution's approximate normality validated 
this approach, though categorical predictors dominated performance.

**Classification Trees:**
- Unpruned Accuracy: 93.2%
- Pruned Accuracy: 93.4%
- Best Specificity: 99.3%
- Pruning slightly improved performance while reducing model complexity. 
Trees excel at capturing non-linear relationships and interactions, though 
they exhibited the most conservative bias toward predicting deportations.

**Models Excluded:**
- Quadratic Discriminant Analysis (QDA) failed due to rank deficiency in 
the covariance matrices, likely caused by high-cardinality categorical 
predictors with sparse combinations
- k-Nearest Neighbors (kNN) was computationally prohibitive given the 
dataset size (367k observations).

### Model Selection Insights
The McNemar's test comparing Random Forest and Logistic Regression revealed
whether their performance differences were statistically significant. 
If significant (p < 0.05), Random Forest's superiority is confirmed; if not, the simpler logistic model may be preferable for interpretability.

**Conservative Prediction Bias:** All models exhibited higher false positive 
rates than false negative rates.

### Limitations and Considerations

**Data Limitations:**
- 15% of observations had unknown apprehension states, potentially introducing
bias if missingness is non-random.
- High-cardinality variables (citizenship country, case category) required 
grouping rare categories, potentially losing granular information
- Single numeric predictor (age) limited the benefit of methods requiring
continuous features.
**Temporal Considerations:**
- Data spans 2023-2025, capturing recent policy environments but limiting 
historical perspective.
- Model performance may degrade if immigration policies shift significantly 
after this period.

## 6.2 Final Remarks

This project successfully demonstrated the application of multiple 
classification techniques to a complex, real-world policy problem. By achieving
93%+ accuracy across diverse modeling approaches, we confirmed that immigration 
case outcomes are highly predictable based on observable case and individual characteristics.
The dominance of case category and criminality status as predictors reveals that deportation decisions are primarily driven by administrative processing pathways
and criminal history rather than demographic factors like age, gender, or even citizenship country. This finding has important implications for both policy 
design and individual case advocacy.

**Critical points:**
It is crucial to emphasize that these conclusions are based on a specific
dataset that may not provide a complete or objective representation of ICE 
enforcement activities. 
Several concerning patterns suggest potential data collection biases or 
selective reporting:

**Anomalous Age-Criminality Patterns:**
Our analysis revealed individuals classified with criminal records at 
implausibly young ages, including minors. This raises questions about data 
quality, classification accuracy, or the definitions used for "criminal records"
(which might include juvenile offenses, family members' records, or data entry errors).

**Potential Selection Bias:**
The dataset represents only cases that resulted in formal ICE processing and documentation. Critically, it likely excludes:
- Individuals apprehended but immediately released without formal case 
processing.
- Cases settled through alternative mechanisms not captured in this 
administrative system.
- Enforcement actions that occurred but were not formally documented.
- Detentions at facilities that do not contribute to this dataset.

---

**Project completed by:** Dmytro Vaskov, Mariia Salivon  
**Course:** Big Data and Uncertainty Quantification  
**University of Vienna**  
**Date:** February 2026


